---
title: "clean_covariates_KC"
author: "Kevin Ngo"
date: "2025-05-19"
output: pdf_document
---


```{r}
library(tidyverse)
```

```{r}
# set data path
DATA_PATH <- "C:/Users/kevoi/OneDrive/Desktop/School/Year 3/POLI SCI 200C/final_project/data"
```


## Aggregate age demographics (20 - 34)

These are the relevant columns we need for 2013-2016

Total population -> Total!!Estimate!!Total population  
People 20 to 24 -> Total!!Estimate!!AGE!!20 to 24 years  
People 25 to 29 -> Total!!Estimate!!AGE!!25 to 29 years  
People 30 to 34 -> Total!!Estimate!!AGE!!30 to 34 years  

These are the relevant columns we need for 2016-2023

Total population -> Total!!Estimate!!Total population  
Total male population -> Estimate!!Total!!Total population  
People 20 to 24 -> Estimate!!Total!!Total population!!AGE!!20 to 24 years
People 25 to 29 -> Estimate!!Total!!Total population!!AGE!!25 to 29 years
People 30 to 34 -> Estimate!!Total!!Total population!!AGE!!30 to 34 years


We will create the following covariates:  

- total_pop: total population  
- perc_age_20_34: percent of populate that is age 20-34

Note: from 2017-2023, the census data does not give the % of people in each age bucket, so we have to manually calculate this

```{r}
AGE_PATH <- file.path(DATA_PATH, "KC_age")
selected_cols_2013_2016 <- c("Total!!Estimate!!Total population", 
                   "Total!!Estimate!!AGE!!20 to 24 years", "Total!!Estimate!!AGE!!25 to 29 years",
                   "Total!!Estimate!!AGE!!30 to 34 years")
selected_cols_2017_2023 <- c("Estimate!!Total!!Total population", 
                   "Estimate!!Total!!Total population!!AGE!!20 to 24 years", 
                   "Estimate!!Total!!Total population!!AGE!!25 to 29 years",
                   "Estimate!!Total!!Total population!!AGE!!30 to 34 years")
covariate_cols <- c("year", "total_pop", "perc_age_20_34")

# create a dataframe to store values for each year
age_df <- data.frame()
```



```{r}
# List all relevant files
files <- list.files(AGE_PATH, pattern = "^ACSST5Y\\d{4}\\.S0101-Data\\.csv$", full.names = TRUE)

# Loop over files
for (file in files) {
  # Extract year from filename
  year <- str_extract(file, "\\d{4}") |> as.numeric()
  
  cat("Reading ", year, " data\n")
  
  # Read and process file
  data <- read_csv(file, skip = 1, show_col_types = FALSE) |> suppressMessages()
    # Since col names of datasets from 2013-2016 differ from 2017-2023, we need this logic to handle diff col names
    if (year <= 2016) { 
      data <- data |> 
        select(all_of(selected_cols_2013_2016)) |> 
        rename(
          total_pop = "Total!!Estimate!!Total population",
        ) |> 
        mutate(
          perc_age_20_34 = (`Total!!Estimate!!AGE!!20 to 24 years` + 
                            `Total!!Estimate!!AGE!!25 to 29 years` + 
                            `Total!!Estimate!!AGE!!30 to 34 years`) / 100,
          year = year
        ) |> 
        select(all_of(covariate_cols))
    } else { 
      data <- data |> 
        select(all_of(selected_cols_2017_2023)) |> 
          rename(
            total_pop = "Estimate!!Total!!Total population",
          ) |> 
          mutate(
            perc_age_20_34 = (`Estimate!!Total!!Total population!!AGE!!20 to 24 years` + 
                              `Estimate!!Total!!Total population!!AGE!!25 to 29 years` + 
                              `Estimate!!Total!!Total population!!AGE!!30 to 34 years`) / total_pop,
            year = year
          ) |> 
          select(all_of(covariate_cols))
    }
    
  # Append to master dataframe
  age_df <- bind_rows(age_df, data)
}
```

```{r}
age_df
```

## Aggregate employment demographics

These are the relevant columns we need

Population 16 years and older <- Estimate!!EMPLOYMENT STATUS!!Population 16 years and over    
Unemployed <- Estimate!!EMPLOYMENT STATUS!!Population 16 years and over!!In labor force!!Civilian labor force!!Unemployed  

We will create the following covariates:  
- % unemployed = Unemployed / Population 16 years and older
- pop_16_up = Population 16 years and older

```{r}
EMPLOYMENT_PATH <- file.path(DATA_PATH, "KC_employment")
selected_cols <- c("Estimate!!EMPLOYMENT STATUS!!Population 16 years and over", 
                   "Estimate!!EMPLOYMENT STATUS!!Population 16 years and over!!In labor force!!Civilian labor force!!Unemployed")
covariate_cols <- c("total_pop_16_up", "perc_unemployed", "year")

# create a dataframe to store values for each year
emp_df <- data.frame()
```

```{r}
# List all relevant files
files <- list.files(EMPLOYMENT_PATH, pattern = "^ACSDP5Y\\d{4}\\.DP03-Data\\.csv$", full.names = TRUE)

# Loop over files
for (file in files) {
  # Extract year from filename
  year <- str_extract(file, "\\d{4}") |> as.numeric()
  
  cat("Reading ", year, " data\n")
  
  # Read and process file
  data <- read_csv(file, skip = 1, show_col_types = FALSE) |> suppressMessages()
    data <- data |> 
      select(all_of(selected_cols)) |> 
      rename(
        total_pop_16_up = "Estimate!!EMPLOYMENT STATUS!!Population 16 years and over",
      ) |> 
      mutate(
        perc_unemployed = `Estimate!!EMPLOYMENT STATUS!!Population 16 years and over!!In labor force!!Civilian labor force!!Unemployed` / total_pop_16_up,
        year = year
      ) |> 
      select(all_of(covariate_cols))
  
  # Append to master dataframe
  emp_df <- bind_rows(emp_df, data)
}
```

```{r}
emp_df
```

## Aggregating gender data

These are the relevant columns we need

total_pop <- Estimate!!SEX AND AGE!!Total population  
total_male <- Male!!Estimate!!Total population

We will create the following covariates:

- perc_male = % of total population that is male

```{r}
SEX_PATH <- file.path(DATA_PATH, "KC_sex")
selected_cols_2013_2016 <- c("Estimate!!SEX AND AGE!!Total population", 
                   "Percent!!SEX AND AGE!!Total population!!Male")
selected_cols_2017_2023 <- c("Estimate!!SEX AND AGE!!Total population", 
                   "Estimate!!SEX AND AGE!!Total population!!Male")
covariate_cols <- c("year", "perc_male")

# create a dataframe to store values for each year
sex_df <- data.frame()
```

```{r}
# List all relevant files
files <- list.files(SEX_PATH, pattern = "^ACSDP5Y\\d{4}\\.DP05-Data\\.csv$", full.names = TRUE)

# Loop over files
for (file in files) {
  # Extract year from filename
  year <- str_extract(file, "\\d{4}") |> as.numeric()
  
  cat("Reading ", year, " data\n")
  
  # Read and process file
  data <- read_csv(file, skip = 1, show_col_types = FALSE) |> suppressMessages()
  
  # Since col names of datasets from 2013-2016 differ from 2017-2023, we need this logic to handle diff col names
  if (year <= 2016) { 
    data <- data |> 
    select(all_of(selected_cols_2013_2016)) |> 
    rename(
      total_pop = "Estimate!!SEX AND AGE!!Total population",
    ) |> 
    rename(perc_male = "Percent!!SEX AND AGE!!Total population!!Male") |> 
    mutate(perc_male = perc_male / 100,
           year = year) |> 
    select(all_of(covariate_cols))
  } else { # from 2017 onwards, the percentage of males isn't given, so we calculate it from scratch
     data <- data |> 
      select(all_of(selected_cols_2017_2023)) |> 
      rename(
        total_pop = "Estimate!!SEX AND AGE!!Total population",
      ) |> 
      rename(total_male = `Estimate!!SEX AND AGE!!Total population!!Male`) |> 
      mutate(perc_male = total_male / total_pop,
              year = year) |> 
      select(all_of(covariate_cols))
  }
    
  # Append to master dataframe
  sex_df <- bind_rows(sex_df, data)
}
```
```{r}
sex_df
```

## Merge all dataframes

```{r}
kc_covariates <- inner_join(age_df, emp_df, by = "year")
kc_covariates <- inner_join(kc_covariates, sex_df, by = "year")
kc_covariates
```
```{r}
write_csv(kc_covariates, "kc_covariates.csv")
```








